%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgments}{acknowledgments}

\begin{flushright}{\slshape
    I would contend at all costs both in word and deed\\
    as far as I could that we will be better men, braver\\
    and less idle, if we believe that one must search for \\
    the things one does not know, rather than if we believe \\
    that it is not possible to find out what we do not know \\
    and that we must not look for it.} \\ \medskip
    --- Plato
\end{flushright}



\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgments}

I am deeply indebted to my advisor, Dr.\ Ryan Tully-Doyle, without his firm
push into the world of free analysis this thesis would have never been possible.
I could not have undertaken this journey without Dr.\ Matthew White and Dr.\
Linda Patton. \{A sentence about them\}. To the countless educators who have given their unrelenting
support to my studies, \{the rest of this sentence\}.

\bigskip
I would also like to extend my deepest gratitude to my family. I am who I am
because of your influence---those of you who
have not yet made it to my arm, know that your flower lives in my heart.
I cannot begin to express my thanks to my fellow graduate students. To Karl and
Caroline, thank you for welcoming me into your little family with such open
arms. To Kelsey, for the many late night McDonalds runs. And to Klig, both for
the sandwiches and the conversations.
Finally, I would like to thank Christian Leonard---his contributions to my life
are immeasurable, but by far the greatest is introducing me to H\"aagen-Dazs
Vanilla Swiss Almond.

\endgroup
